---
title: "Data Cleaning and Analysis"
author: "Jesse Richardville"
date: "2023-01-05"
output: html_document
---
# Document Summary
Clean and transform the merged Cyclistic data (all_data) file for analysis using R
​
Raw data has been collected by Motivate International Inc, the company operating Divvy - a bike share service across Chicago and Evanston. The license to use this public dataset can be found [here.](https://www.divvybikes.com/data-license-agreement)
​
## Upload the merged dataset from the working directory
```{r setup, include=FALSE}
library(readr)
all_bike_trip_merged <- read_csv("R docs/Working Directory/Cyclistic_case_study/data/raw/dirty_bike_data/all_bike_trip_merged.csv")
View(all_bike_trip_merged)
all_data <- all_bike_trip_merged

```

### Install appropriate packages
```{r setup, include=FALSE}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("Magrittr")
library(tidyverse)
library(lubridate)
library(Magrittr)

# calculate ride length and add as a new column, ride_time
all_data$ride_time <- difftime(
  all_data$ended_at, 
  all_data$started_at,
  units = "secs"
) 

# convert ride_time to hms for readability
library(data.table)
all_data <- all_data %>% 
  mutate(ride_hms = as.ITime(hms::hms(seconds_to_period(ride_time)))
  )

# reposition columns in table for order
all_data <- all_data %>% relocate(ride_time, .after = ended_at)
all_data <- all_data %>% relocate(ride_hms, .after = ride_time)

# remove rides <60 seconds for outliers/not real rides
all_data <- arrange(all_data, -ride_time)
all_data <- arrange(all_data, ride_time)
all_data <- all_data %>%
  filter(!(ride_time < 60)
  )

# remove rides >12 hours (43200 seconds) for outliers/not real rides
all_data <- all_data %>%
  filter(!(ride_time > 43200)
  )

# remove rides missing data (without start_station_name or end_station_name)
all_data <- all_data %>%
  filter(
    !(is.na(start_station_name) |
        start_station_name == "")
  ) %>% 
  filter(
    !(is.na(end_station_name) |
        end_station_name == "")
  )

# check for abnormal station names - rides with ALL CAPS or all lower 
UPPER_lower_station_name_check <- all_data %>%
  filter(
    str_detect(start_station_name, "[:upper:]")
    & !str_detect(start_station_name,"[:lower:]")
  ) %>%
  group_by(
    start_station_name
  ) %>%
  count(
    start_station_name
  )
# one station appeared 4 times with WEST CHI-WATSON, determined legitimate

# check for test data by star_station_name
library(stringr)
Test <- all_data[str_detect(all_data$start_station_name, "Test"), ]
all_data <- all_data %>%
  filter(
    !(str_detect(start_station_name, "Test"))
  )
# removed 1 result containing "Test" in start_station_name

# check for duplicates
install.packages("janitor")
library(janitor)
duplicates <- get_dupes(all_data, ride_id)   
#No variable names specified - using all columns.

# removed unnecessary column (1) from table
all_data <- all_data[, -c(1)]

# check for anomalies by running a summary of the dataset
install.packages("psych")
library(psych) 
# create summary table
anomalies <- describe(all_data)
# no abnormalities, except there were 7 cases where the end_lat and end_lng 
# were 0 but they had an end_station_name and end_station_id (were removed)

# check bike types and when they were available
unique(all_data$rideable_type)
[1] "classic_bike"  "electric_bike" "docked_bike"  
rideable_type_info <-all_data %>%
  mutate(
    year = year(started_at), 
    month = month(started_at)
  ) %>%
  group_by(
    month, 
    year
  ) %>%
  select(
    rideable_type, 
    member_casual,
    month, 
    year
  ) %>%
  count(
    rideable_type, 
    member_casual
  )
# After filtering the raw data, all data for docked_bike is for casual riders.
# Docked bikes are not used by member riders, but I'm not sure why.


# check station names for review
station_names <- all_data %>%
  mutate(
    year = year(started_at), 
    month = month(started_at)
  ) %>%
  group_by(
    month, 
    year
  ) %>%
  select(
    start_station_name, 
    month, 
    year
  ) %>%  
  count(start_station_name)

# create columns of data by 'day', 'day of week', 'month', and 'year' for future analysis/viz
# Day
all_data$day <- format(
  all_data$started_at, 
  "%d"
  )
# Day of week 
all_data$day_of_week <- format(
  all_data$started_at, 
  "%A"
  )
# Month 
all_data$month <- format(
  all_data$started_at, 
  "%m"
  )
# Year 
all_data$year <- format(
  all_data$started_at, 
  "%Y"
  )
# Time of Day, HH:MM:SS
all_data$ToD <- format(
  all_data$started_at, 
  "%H:%M:%S"
  )

# statistical analysis by user type
summary(all_data$ride_time)
Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
60     376     649    1012    1156   43147 

aggregate(all_data$ride_time ~ all_data$member_casual, FUN = mean)
all_data$member_casual all_data$ride_time
1                 casual          1400.7239
2                 member           750.5077
aggregate(all_data$ride_time ~ all_data$member_casual, FUN = median)
all_data$member_casual all_data$ride_time
1                 casual                848
2                 member                549
aggregate(all_data$ride_time ~ all_data$member_casual, FUN = max)
all_data$member_casual all_data$ride_time
1                 casual              43147
2                 member              43020
aggregate(all_data$ride_time ~ all_data$member_casual, FUN = min)
all_data$member_casual all_data$ride_time
1                 casual                 60
2                 member                 60

# create tables for visuals to be used in viz
# map tables for visual
map_viz_start <- all_data %>%
  select(
    start_station_name, 
    start_lat, 
    start_lng
  ) %>%
  group_by(
    start_station_name
  ) %>%
  mutate(
    numtrips = n()
  ) %>%
  distinct(
    start_station_name,  
    .keep_all = TRUE
  )

map_viz_end <- all_data %>%
  select(
    end_station_name, 
    end_lat, 
    end_lng
  ) %>%
  group_by(
    end_station_name
  ) %>%
  mutate(
    numtrips = n()
  ) %>%
  distinct(
    end_station_name,  
    .keep_all = TRUE
  )

```